---
title: "vicspatial for Beginners"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{vicspatial for Beginners}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
Modifications Copyright 2020 Rachel Swain and Justin Cally
Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->

```{r, echo = FALSE}
# Connected only if the geoserver can actually serve a filtered query. The base
# URL can be reachable while filtered WFS requests are rejected (e.g. CI /
# datacentre IPs receive HTTP 400); in that case skip the live chunks.
geoserver_connected <- tryCatch({
  n <- vicspatial::feature_hits(dplyr::filter(vicspatial::vicmap_query("open-data-platform:hy_watercourse"), hierarchy == "L"))
  is.numeric(n) && !is.na(n) && n > 0
}, error = function(e) FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>",
  eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3),
  purl = FALSE
)
```

```{r, include=FALSE, echo=FALSE, eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
# LIBRARIES -----
library(vicspatial)
library(dplyr)
library(sf)
```

## Who is this tutorial for?

The purpose of this tutorial is:

 - To introduce Vicmap
 - To explain the advantages of vicspatial in more detail
 - To provide simple examples of querying, visualising and working with Vicmap spatial data
 
The end goal of this tutorial is for R users with minimal spatial experience to be able to search for, download and visualise Vicmap data. While this tutorial is aimed at spatial beginners, it can still be used as a reference for more experienced users.


## What is vicspatial?

vicspatial is the R package that connects to the Victorian Government's catalogue of spatial datasets. The catalogue features `r try(nrow(vicspatial::listLayers()))` datasets across land, property, infrastructure and environment and is the most authoritative suite of spatial data in Victoria. 

### Accessing vicspatial datasets

This data catalogue is freely available to the public and can be accessed via several methods. For users of R, the fastest way to access up to date vicspatial datasets is to utilise the Web Feature Service (WFS). WFS is a standardised interface to request geographic information, regardless of the platform on which it is stored. 

WFS requires a URL that contains the instructions for the query and is written in WFS specific terminology. This of course requires understanding of the WFS terminology and the time-consuming process of manually building the URL string. 


## Enter, vicspatial.

#### WFS, without the fuss!

vicspatial simplifies WFS queries by taking user supplied keywords and automatically building the correctly formatted URL string.  
The functions in vicspatial therefore provide a much faster and more robust method for acquiring data.

#### Lazy evaluation

vicspatial uses *lazy evaluation*,  in which we query but do not collect the data from the WFS database. This is called making a 'Promise'. A promise contains the dataset information but not the data itself. This promise can be filtered for a subset of the data before collecting. Using lazy evaluation we can collect the desired subset, instead of the entire dataset.

## Worked Example - Melbourne's Open Space Network

*Note: A previous version of this tutorial used swooping birds as the example, but due to the migration of platforms that data was not available at time of writing*

To demonstrate vicspatial's features, we will download and explore a dataset from the data catalogue. The data is a spatial dataset of Melbourne's open space network.

<!-- Australian spring is famous for the presence of swooping birds. Certain species of birds swoop during the breeding season to protect their nest from potential predators. Swooping birds in Victoria include the masked lapwing (plover), butcherbird, magpie-lark, noisy miner and perhaps most famously, the magpie. The Vicmap catalogue has a dataset of 'swoop' locations, recorded by members of the public.  -->

### Searching for data 
First, we need to search for the dataset. We can view all layers, or do a keyword search. Use `ignore.case = TRUE` for a case-insensitive search. `listLayers()` by default now provides a column of 'Abstract' and 'metadataID'. You can remove these columns by setting the abstract argument to FALSE (e.g. `listLayers(abstract = FALSE)`)

```{r,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}

#All layers
all_layers <- listLayers() 

# Case insensitive keyword search
search <- listLayers(pattern = "veac", ignore.case = T) 

```

Viewing our search results shows us the name of the dataset and its description.

```{r echo=FALSE, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
library(knitr)
library(kableExtra)
search %>% kable() #%>% kable_styling("bordered")
```

Once we have the name of the dataset we would like to download, in this case *'open-data-platform:veac_metro_open_space'*, we can query the Vicmap database. 

If we are still not sure from the name and description that this is the right dataset, we can look at the promise to get a snippet of the dataset contents. 

```{r, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
vicmap_query(layer = "open-data-platform:veac_metro_open_space") 
```

We can see that the dataset contains the route numbers and names. At a glance this data should be adequate. If you want more extensive information of the data you can use `get_metadata()` to download a list of (i) metadata about the data, (ii) a data dictionary (*work-in-progress*) and (iii) a link to the metadata url. 

```{r, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
metadata <- vicmap_query(layer = "open-data-platform:veac_metro_open_space") %>%
  get_metadata() %>%
  .[[1]]

kbl(metadata) %>%
  kable_styling()
```

### Downloading the data

This isn't a very large dataset, but if we are using this data in an interactive report or application, waiting to download the full dataset may be impractical. 

Using lazy evaluation, we can use pipes to filter and subset the data, so that we only collect the desired subset. For example, let's just look at the first 10 rows.

```{r, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
query <- vicmap_query(layer = "open-data-platform:veac_metro_open_space") %>% 
  head(10) %>% collect()

query

```

### Subsetting data

Now let's only look at the open space data. Note that vicspatial datasets will retain `id` and `geometry` columns even if not selected in the vicspatial promise. As this a simple features (sf) dataset, the geometry corresponding to the features does not need to be selected as it is always attached. The geometry column can only be removed intentionally, for example, by using the `st_drop_geometry()` function in the `sf` package.

```{r, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
veac_select <- vicmap_query(layer = "open-data-platform:veac_metro_open_space") %>% 
  filter(dominant_lga_name == "DAREBIN") %>% 
  filter(ha > 0.1) %>% 
  select(id, ownername_summary, ha, os_group, dominant_lga_name, name) %>%
  collect()

 veac_select %>% head(5) %>% 
  select(-id) %>% #condense table for viewing
  kable() %>% 
  kable_styling()
```

**By filtering the `Vicmap_promise` before collecting, we have downloaded only a subset of rows of data.**


### Quick visualisation of spatial data with the `mapview` package

Often the first thing we want to do is see the data on a map. The simplest and quickest way to visualise spatial data is to use the `mapview` package:

```{r, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
library(mapview)
mapviewOptions(fgb = FALSE)

# Because the dataset is not big we can download all the data as 'veac_area'
veac_area <- vicmap_query(layer = "open-data-platform:veac_metro_open_space") %>% 
  collect() 

# plot data
mapview::mapview(veac_area)
```

While the output is basic, it is very simple to achieve and lines can be clicked on to obtain the details from the other fields.


### Visualisation with `leaflet`

For more customization, you might want to consider plotting your map in `leaflet`. Leaflet is JavaScript library that allows users to produce interactive maps that work efficiently across most desktop and mobile platforms. Using the leaflet package in R, we have the flexibility of plotting more complex map products.

We will start by plotting the most basic map in leaflet. Instead of using a single function like `mapview()`, in leaflet we need to add each component to the map. At a minimum we specify the leaflet object, add a basemap and add the markers. Note: differing to mapview, leaflet does not include marker labels by default, these must be added with the `popup` parameter.

```{r, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
library(leaflet)

# Create a palette
pal <- colorFactor("Set1", levels = veac_area$os_group) #define a colour palette

veac_area %>%
  leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron) %>% #add third party base map
  addPolygons(
    color = ~pal(os_group),
    weight = 2,
    stroke = 0.5, #removes outline
    fillOpacity = 0.8,
    popup = paste0("<b>Group.: </b>", veac_area$os_group, "<br>", #format the popup with html tags
                   "<b>LGA Name: </b>", veac_area$dominant_lga_name, "<br>",
                   "<b>Name: </b>", veac_area$name)
  )
```

### Spatial filtering

Let's say we want to know about the VEAC area only in our LGA (e.g. Darebin). We can apply a spatial filter to our dataset so that only the areas in our suburb are shown. Where do we get the spatial data for our LGA? From the Vicmap catalogue of course!

## Using the vicspatial Geometric Filters

The vicspatial package offers tools for geometric filtering that avoid the need for sf package. The WFS Geoserver on which Vicmap is based supports several geometric filters - see the full list [here](https://justincally.github.io/vicspatial/articles/query_vicmap.html#geometric-filters).

These geometric filters allow us to perform basic spatial manipulation while retaining the benefits of lazy evaluation. So as before, if we want to filter our data before downloading, we can do so with more complex spatial operations. This is particularly useful when presenting vicspatial data in an interactive map, as it reduces the time for downloading data between re-rendering the map.

To get the areas only in Darebin we can filter with `INTERSECTS()`.

```{r, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}
# areas in Darebin
darebin <- vicmap_query(layer = "open-data-platform:lga_polygon") %>%
  select(lga_name) %>% 
  filter(lga_name == "DAREBIN") %>%
  collect() %>%
  st_make_valid() # magic fix for some spatial data

darebin_veac <- vicmap_query(layer = "open-data-platform:veac_metro_open_space") %>%
  filter(INTERSECTS(darebin)) %>%
  collect()

mapview::mapview(darebin) + mapview::mapview(darebin_veac, alpha.regions = 0.3, col.regions = "orange")

```
And finally, the veac open spaces on our cycling route. To find spaces within 30m we use the `INTERSECTS()` function and the cycling route with an added 30m buffer. Note: vicspatial geometric filters will be simplified, thus if precise intersections are desired, it is advised to do a cleanup of the filtered data once collected.

```{r, warning=FALSE, message=FALSE,eval = all(geoserver_connected, !testthat:::on_cran(), sf::sf_extSoftVersion()[["GDAL"]] > 3)}

route_line <- sf::st_read(system.file("shapes/cycle_route.geojson", package="vicspatial"), quiet = F) %>% 
  sf::st_transform(4283)

# Condense line object
route_poly <- route_line %>% 
  sf::st_transform(3111) %>% 
  sf::st_buffer(30) %>% 
  sf::st_cast("POLYGON") %>% 
  sf::st_transform(4283)

route_intersection <- vicmap_query(layer = "open-data-platform:veac_metro_open_space") %>%
  filter(INTERSECTS(route_poly)) %>% 
  collect() 


mapview::mapview(route_poly) + mapview::mapview(route_intersection, color = "red", lwd = 5)

```


## Debrief

After completing this tutorial you should have a basic understanding of plotting spatial data and performing some basic spatial operations. This tutorial is a quick guide and designed only to get you started. For further learning, see the following resources:

*  [Simple Features for R](https://r-spatial.github.io/sf/)
*  [Leaflet for R](http://rstudio.github.io/leaflet/)
*  [Introduction to Spatial Data Science in R](https://rspatial.org/intr/1-introduction.html)
