---
title: "How to Query Vicmap Data"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to Query Vicmap Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<!--
Copyright 2020 Province of Justin Cally

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->

```{r, echo = FALSE}
# Connected only if the geoserver can actually serve a filtered query. The base
# URL can be reachable while filtered WFS requests are rejected (e.g. CI /
# datacentre IPs receive HTTP 400); in that case skip the live chunks.
geoserver_connected <- tryCatch({
  n <- vicspatial::feature_hits(dplyr::filter(vicspatial::vicmap_query("open-data-platform:hy_watercourse"), hierarchy == "L"))
  is.numeric(n) && !is.na(n) && n > 0
}, error = function(e) FALSE)

knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>", 
  eval = all(geoserver_connected, !testthat:::on_cran()),
  purl = FALSE
)
```

```{r setup}
library(vicspatial)
library(sf)
library(leaflet)


#check sf installation
sf::sf_extSoftVersion()
```

# How to query data

## Searching for data  

In order to begin a query of the WFS server a spatial layer must be selected. To know which layers are available use the `listLayers()` function, which will return ~ 690 layers to choose from. 

```{r}
available_layers <- listLayers()

head(available_layers, 10)
```


## Vicmap promise

vicspatial introduces a new class called `vicmap_promise`, which is an extension to the `httr::url` class. Essentially this object is how the vicmap query is stored before data is collected. That is to say `vicmap_promise` is essentially a promise of what data will be retrieved. 

In order to generate a new promise the `vicmap_query` function can be used to select the layer. The promise prints a sample of the data (max = 6 rows) as well as the dimensions (nrow and ncol).  

```{r query}
# query the watercourse layer
vicmap_query(layer = "open-data-platform:hy_watercourse")
```

## Adding arguments to the query  

The `vicmap_promise` object can be easily added to through piping in of additional functions (e.g. `head()`, `filter()` and `select()`). 

The resulting query can be displayed using the `show_query()` function, which will list the WFS parameters. 

```{r query_arguments}
vicmap_query(layer = "open-data-platform:hy_watercourse") %>%
  head(50) %>% #return only 50 rows
  filter(hierarchy == "L") %>% # filter the column 'HIERACHY' to values of 'L'
  select(hierarchy, pfi) %>% # select columns 'HIERARCHY' and 'PFI'
  show_query()
```

In order to return a spatial data.frame object (`sf`) `collect()` must be used.

```{r collect_query}
watercourse_data <- vicmap_query(layer = "open-data-platform:hy_watercourse") %>%
  head(50) %>% #return only 50 rows
  filter(hierarchy == "L") %>% # filter the column 'HIERACHY' to values of 'L'
  select(hierarchy, pfi) %>% # select columns 'HIERARCHY' and 'PFI'
  collect()

str(watercourse_data)
```


## Geometric filters

vicspatial translates numerous geometric filter functions available in the Victorian Government's WFS Geoserver supports numerous [geometric filters](https://docs.geoserver.org/stable/en/user/tutorials/cql/cql_tutorial.html#geometric-filters):  

+ `EQUALS`   
+ `DISJOINT`   
+ `INTERSECTS`   
+ `TOUCHES`   
+ `CROSSES`   
+ `WITHIN`   
+ `CONTAINS` 
+ `OVERLAPS`   
+ `DWITHIN`   
+ `BEYOND`  
+ `BBOX`  

These filters can be used within the `filter()` function by providing them an object of class `sf/sfc/sfg/bbox`. Below is a leaflet map with the melbourne rail network being read in with the use of three different types of filter functions: `INTERSECTS()`, `BBOX()` and `DWITHIN()`.

```{r filter_spatial, warning = FALSE, message=FALSE, out.width="100%"}
#### Return objects that intersect melbourne ####
# Read in an example shape to restrict our query to using geometric filtering
melbourne <- sf::st_read(system.file("shapes/melbourne.geojson", package="vicspatial"), quiet = F) %>% 
  sf::st_transform(4283)

# Return data that intersects melbourne
rail_intersects <- vicmap_query(layer = "open-data-platform:tr_rail") %>% # layer to query
  filter(INTERSECTS(melbourne)) %>% # more advanced geometric filter
  collect()

melbourne_bbox <- sf::st_bbox(melbourne)

rail_bbox <- vicmap_query(layer = "open-data-platform:tr_rail") %>%
  filter(BBOX(melbourne_bbox)) %>%
  collect()

melbourne_centroid <- sf::st_centroid(melbourne)

rail_dwithin <- vicmap_query(layer = "open-data-platform:tr_rail") %>%
  filter(DWITHIN(melbourne_centroid, distance = 10000, units = "meters")) %>%
  collect()

leaflet(width = "100%") %>%
  addProviderTiles("CartoDB.Positron") %>%
  addPolygons(data = melbourne, color = "grey", group = "Melbourne polygon") %>%
  addPolygons(data = sf::st_bbox(melbourne) %>% st_as_sfc(), color = "black", group = "Melbourne bbox") %>%
  addPolylines(data = rail_intersects, color = "Red", group = "INTERSECTS") %>% 
  addPolylines(data = rail_bbox, color = "Blue", group = "BBOX") %>%
  addPolylines(data = rail_dwithin, color = "Green", group = "DWITHIN") %>%
  addLayersControl(baseGroups = c("Melbourne polygon", "Melbourne bbox"), 
                   overlayGroups = c("INTERSECTS", "BBOX", "DWITHIN")) %>%
  hideGroup(c("BBOX", "DWITHIN"))
```


